Semantics for the $ \lambda $-calculus

### 2024-03-05 report/chapters/previous-work.tex
Remark upon the importance of a reflexive object

\TODO Remark upon the significance of having a reflexive object

We can then wonder whether this construction is part of a correspondence between untyped $ \lambda $-calculi and some class of categories. This question finds a partial answer in the following theorem, originally proven in a very syntactical way by Dana Scott (see \cite{curry}, page 418).
\begin{lemma}
  We can obtain every untyped $ \lambda $-calculus as the endomorphism theory of some object in some category.
\end{lemma}
\begin{proof}
  Let $ L $ be a $ \lambda $-theory. First of all, for $ a_1, a_2: L_0 $, we define
  \begin{align*}
    a_1 \circ a_2 &= \lambda x_1, a_1 (a_2 x_1);\\
    (a_1, a_2) &= \lambda x_1, x_1 a_1 a_2;\\
    \langle a_1, a_2 \rangle &= \lambda x_1, (a_1 x_1, a_2 x_1);\\
    \pi_i &= \lambda x_1, x_1 (\lambda x_2 x_3, x_{i + 1}).
  \end{align*}
  Although, actually, since every one of these starts with a $ \lambda $-abstraction, we would need to lift the constants $ a_i $ to $ \iota_{0, 1}(a_i): L_1 $ to make the definitions above typecheck.

>>>>

\begin{remark}
  To construct a \textit{simply typed} $ \lambda $-calculus from a category, we just need a cartesian closed category. In a simply typed $ \lambda $-calculus, there is a lot of restriction on which terms we can apply to each other. A term of type $ A \to B $ can only be applied to a term of type $ A $, which gives a term of type $ B $. In particular, a term can be applied only finitely many times to other terms, and every time, the result has a different type.

  On the other hand, for an \textit{untyped} $ \lambda $-calculus, we need a cartesian closed category with a `reflexive object'. This is because in the untyped $ \lambda $-calculus, we can apply arbitrary terms to each other. For example, we can apply the term $ (\lambda x_1, x_1 x_1) $ to itself, which would not be typable in the simply typed $ \lambda $-calculus. Suppose that we have a category $ C $ and an object $ U $ such that the morphisms $ C(U^n, U) $ give the untype $ \lambda $-terms in $ n $ free variables, for all $ n $. Now, given two terms $ f, g : C(U^n, U) $, for the application $ f g $, we need to consider $ f $ as a morphism in $ C(U^n, U^U) $. We can do this by postcomposing with a morphism $ \varphi: C(U, U^U) $. On the other hand, if $ n > 0 $, then $ (\lambda x_n, f) $ is a morphism in $ C(U^{n - 1}, U^U) $, but it is a term in $ n - 1 $ free variables, so it should be in $ C(U^{n - 1}, U) $. For this, we postcompose with a morphism $ \psi: C(U^U, U) $. Now, for our untyped $ \lambda $-calculus to have $ \beta $-equality, we need $ \psi \cdot \varphi = \id{U^U} $, which means that the exponential $ U^U $ of $ U $ is a retract of $ U $. This is exactly what it means for $ U $ to be a \iindex{reflexive object}: an object $ U $ in a cartesian closed category, that has a retraction onto its `function space' $ U^U $. Note that if we want our $ \lambda $-theory to also have $ \eta $-equality, the retraction must be an isomorphism.

  Note that $ \SET $ is a cartesian closed category, but that for sets $ X $ and $ Y $, the function space $ X^Y $ has cardinality $ \vert X \vert^{\vert Y \vert} $, and therefore $ U^U $ cannot be a retract of $ U $, unless $ U = \{ \star \} $, in which case we have a very trivial $ \lambda $-calculus: $ \SET(U^n, U) = \{ \star \} $.

  During the 1960s, computer scientists sought for nontrivial examples of reflexive objects, there is a quote by Dana Scott that ``Lambda-calculus has no mathematical models!''\cite{strachey}. However, in 1969, the same Dana Scott discovered that the category of (continuous) lattices with (Scott-)continuous functions between them has a nontrivial reflexive object, with the retraction onto the function spaces being even an isomorphism. Such a reflexive object $ D_\infty $ is obtained by starting with an arbitrary lattice $ D_0 $, iteratively taking $ D_{n + 1} = D_n^{D_n} $, with a retraction (in the `wrong' direction) $ r: D_n^{D_n} \to D_n $, and then passing to the limit $ D_\infty = \lim_{\leftarrow} D_n $ (for the main result, see Theorem 4.4 in \cite{scott-continuous}, page 127).
\end{remark}

Since Lambek showed an equivalence between simply typed $ \lambda $-calculi and cartesian closed categories, and since we have a construction for an untyped $ \lambda $-calculus from a cartesian closed category with a reflexive object, we can wonder whether this construction constitutes contitutes an equivalence between untyped $ \lambda $-calculi and some class of categories. This question finds a partial answer in the following theorem, originally proven in a very syntactical way by Dana Scott (see \cite{curry}, page 418).
\begin{lemma}
  We can obtain every untyped $ \lambda $-calculus as the endomorphism theory of some object in some category.
\end{lemma}
\begin{proof}
  Let $ L $ be a $ \lambda $-theory. First of all, for $ a_1, a_2: L_0 $, we define
  \begin{align*}
    a_1 \circ a_2 &= \lambda x_1, a_1 (a_2 x_1);\\
    (a_1, a_2) &= \lambda x_1, x_1 a_1 a_2;\\
    \langle a_1, a_2 \rangle &= \lambda x_1, (a_1 x_1, a_2 x_1);\\
    \pi_i &= \lambda x_1, x_1 (\lambda x_2 x_3, x_{i + 1}).
  \end{align*}
  Although, actually, since every one of these starts with a $ \lambda $-abstraction, we would need to lift the constants $ a_i $ to $ \iota_{0, 1}(a_i): L_1 $ to make the definitions above typecheck.
